---
title: "K-Means Assignment"
author: "Jean Alvarez"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 3
    theme: cosmo
  revealjs:
    embed-resources: true
    transition: zoom
    toc: true
    toc-title: "Overview"
    scrollable: true
---

## Introduction

The `mtcars` dataset contains fuel consumption and 10 aspects of automobile design and performance for 32 cars from 1973–74. All variables are numeric, making this dataset well-suited for K-means clustering. We will determine the optimal number of clusters using the elbow and silhouette methods, then characterize the resulting groups.

## Load Data and Setup

```{r}
#| label: setup
#| message: false
library(factoextra)
library(cluster)

data(mtcars)
str(mtcars)
summary(mtcars)
```

Since K-means is sensitive to scale, we standardize all variables before clustering.

```{r}
#| label: scale-data
mtcars_scaled <- scale(mtcars)
head(mtcars_scaled)
```

## Determining Optimal k

### Elbow Method

The elbow method plots total within-cluster sum of squares (WSS) against k. We look for the "elbow" where adding more clusters yields diminishing returns.

```{r}
#| label: elbow-method
#| fig-cap: "Elbow Method — Total Within-Cluster Sum of Squares vs. k"
set.seed(42)
fviz_nbclust(mtcars_scaled, kmeans, method = "wss", k.max = 10, nstart = 25) +
  geom_vline(xintercept = 3, linetype = 2, color = "red") +
  labs(title = "Elbow Method for Optimal k",
       subtitle = "Dashed line at k = 3")
```

The curve shows a clear bend at **k = 3**, after which the decrease in WSS slows considerably.

### Silhouette Method

The silhouette method measures how similar each point is to its own cluster versus neighboring clusters. Higher average silhouette width indicates better-defined clusters.

```{r}
#| label: silhouette-method
#| fig-cap: "Average Silhouette Width vs. k"
set.seed(42)
fviz_nbclust(mtcars_scaled, kmeans, method = "silhouette", k.max = 10, nstart = 25) +
  labs(title = "Silhouette Method for Optimal k")
```

The silhouette analysis confirms that **k = 2** gives the highest average silhouette width, but **k = 3** is a close second and provides more granular insight. Balancing parsimony with interpretability, we proceed with **k = 3**.

## K-Means Clustering (k = 3)

```{r}
#| label: kmeans-fit
set.seed(42)
km <- kmeans(mtcars_scaled, centers = 3, nstart = 25)
km
```

## Cluster Visualization

### PCA Biplot

We project the 11-dimensional data onto its first two principal components to visualize cluster separation.

```{r}
#| label: pca-biplot
#| fig-cap: "Clusters projected onto first two principal components"
fviz_cluster(km, data = mtcars_scaled,
             ellipse.type = "convex",
             palette = "jco",
             ggtheme = theme_minimal(),
             main = "K-Means Clusters (k = 3) — PCA Biplot")
```

### Silhouette Plot

```{r}
#| label: silhouette-plot
#| fig-cap: "Silhouette plot for k = 3"
sil <- silhouette(km$cluster, dist(mtcars_scaled))
fviz_silhouette(sil, palette = "jco") +
  labs(title = "Silhouette Plot (k = 3)")
```

## Cluster Profiles

### Summary Statistics by Cluster

```{r}
#| label: cluster-profiles
mtcars$cluster <- factor(km$cluster)

cluster_means <- aggregate(. ~ cluster, data = mtcars, FUN = mean)
cluster_means
```

### Heatmap of Cluster Centers (Standardized)

```{r}
#| label: heatmap
#| fig-cap: "Heatmap of standardized cluster centers"
centers_df <- as.data.frame(km$centers)
centers_df$cluster <- factor(1:3)

centers_long <- tidyr::pivot_longer(centers_df, -cluster,
                                     names_to = "variable",
                                     values_to = "value")

ggplot2::ggplot(centers_long, ggplot2::aes(x = variable, y = cluster, fill = value)) +
  ggplot2::geom_tile(color = "white") +
  ggplot2::scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                                midpoint = 0) +
  ggplot2::labs(title = "Cluster Centers (Standardized)", x = "", y = "Cluster") +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

### Boxplots of Key Variables by Cluster

```{r}
#| label: boxplots
#| fig-cap: "Distribution of key variables across clusters"
#| fig-height: 8
key_vars <- c("mpg", "hp", "wt", "disp", "drat", "qsec")
par(mfrow = c(3, 2), mar = c(4, 4, 2, 1))
for (v in key_vars) {
  boxplot(as.formula(paste(v, "~ cluster")), data = mtcars,
          col = c("#0073C2", "#EFC000", "#868686"),
          main = v, xlab = "Cluster", ylab = v)
}
```

### Cars in Each Cluster

```{r}
#| label: cluster-members
for (i in 1:3) {
  cat("\n--- Cluster", i, "---\n")
  print(rownames(mtcars[mtcars$cluster == i, ]))
}
```

## Cluster Descriptions

```{r}
#| label: description-table
#| echo: false
cluster_means_rounded <- cluster_means
cluster_means_rounded[-1] <- round(cluster_means_rounded[-1], 1)
knitr::kable(cluster_means_rounded,
             caption = "Mean values of each variable by cluster")
```

Based on the cluster profiles above:

- **Cluster 1 — Efficient & Lightweight:** High mpg, low displacement and horsepower, low weight. These are the economy cars (e.g., Toyota Corolla, Honda Civic, Fiat 128). They tend to have lower cylinder counts (4-cyl) and higher rear axle ratios.

- **Cluster 2 — Heavy & Powerful:** Very low mpg, high displacement, high horsepower, and heavy weight. These are the muscle cars and large sedans (e.g., Chrysler Imperial, Cadillac Fleetwood, Lincoln Continental). Predominantly 8-cylinder with automatic transmissions.

- **Cluster 3 — Balanced / Sport:** Moderate mpg, moderate-to-high horsepower, and mid-range weight. These sit between the other two groups and include sportier or mid-size vehicles (e.g., Mazda RX4, Merc 280). They often have better quarter-mile times relative to their power.

## Conclusion

Using both the elbow and silhouette methods, we identified **3 meaningful clusters** in the `mtcars` dataset. The clusters correspond to intuitive groupings: economy cars, heavy/powerful vehicles, and a balanced middle class. The PCA biplot shows good separation, and the silhouette plot confirms reasonable cluster cohesion with few misclassified observations.

