---
title: "K-Means Assignment"
author: "Jean Alvarez"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 3
    theme: cosmo
  revealjs:
    embed-resources: true
    transition: zoom
    toc: true
    toc-title: "Overview"
    scrollable: true
---

## Setup

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(h2o)
library(cluster)
library(GGally)

# Load data
data(mtcars)
```

## Explore the Data

```{r}
head(mtcars)
str(mtcars)
```

All variables in `mtcars` are numeric, so we use all of them for clustering.

```{r}
#| fig-width: 10
#| fig-height: 10
#| message: false

ggpairs(mtcars) +
  theme_bw()
```

## Initialize H2O

```{r}
#| message: false

h2o.init()
h2o.no_progress()

# Upload mtcars to H2O
mtcars_h2o <- as.h2o(mtcars)

# All column names (all numeric)
features <- colnames(mtcars)
```

## Part 1: Elbow and Silhouette Methods

### Elbow Method

```{r}
#| message: false

get_inertia <- function(k, data = mtcars_h2o) {
  km_model <- h2o.kmeans(
    training_frame = data,
    x = features,
    k = k,
    standardize = TRUE,
    seed = 123
  )
  return(h2o.tot_withinss(km_model))
}

k_values <- 1:15

elbow_results <- tibble(
  K = k_values,
  Inertia = map_dbl(k_values, get_inertia)
)
```

```{r}
elbow_results |>
  ggplot(aes(x = K, y = Inertia)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = k_values) +
  labs(
    title = "Elbow Method",
    x = "Number of Clusters (K)",
    y = "Total Within-Cluster Sum of Squares (Inertia)"
  ) +
  theme_bw()
```

The elbow plot shows a notable bend around **K = 3**, where the rate of decrease in inertia begins to flatten. This suggests that 3 clusters capture most of the meaningful structure in the data.

### Silhouette Method

```{r}
#| message: false

get_silhouette <- function(k) {
  km_model <- h2o.kmeans(
    training_frame = mtcars_h2o,
    x = features,
    k = k,
    standardize = TRUE,
    seed = 123
  )

  clusters <- as.vector(h2o.predict(km_model, mtcars_h2o))

  dist_matrix <- mtcars |>
    scale() |>
    dist()

  sil <- silhouette(clusters, dist_matrix)
  return(mean(sil[, 3]))
}

silhouette_results <- tibble(
  K = 2:15,
  Avg_Silhouette = map_dbl(2:15, get_silhouette)
)
```

```{r}
silhouette_results |>
  ggplot(aes(x = K, y = Avg_Silhouette)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = 2:15) +
  labs(
    title = "Silhouette Method",
    x = "Number of Clusters (K)",
    y = "Average Silhouette Score"
  ) +
  theme_bw()
```

The silhouette method confirms **K = 2** has the highest average silhouette score, but **K = 3** also performs well. Combining both methods, **K = 3** provides a good balance between cluster compactness and meaningful separation, so we proceed with K = 3.

## Part 2: Cluster Analysis with K = 3

### Fit Final Model

```{r}
#| message: false

final_model <- h2o.kmeans(
  training_frame = mtcars_h2o,
  x = features,
  k = 3,
  standardize = TRUE,
  seed = 123
)

h2o.tot_withinss(final_model)
```

### Assign Clusters

```{r}
groups <- as.vector(h2o.predict(final_model, mtcars_h2o))

mtcars_clustered <- mtcars |>
  mutate(cluster = as.factor(groups))
```

### Cluster Membership

```{r}
mtcars_clustered |>
  group_by(cluster) |>
  summarize(n = n())
```

### Cars in Each Cluster

```{r}
mtcars_clustered |>
  rownames_to_column("car") |>
  group_by(cluster) |>
  summarize(cars = paste(car, collapse = ", "))
```

### Cluster Profiles (Mean Values)

```{r}
cluster_means <- mtcars_clustered |>
  group_by(cluster) |>
  summarize(across(where(is.numeric), mean))

cluster_means
```

### Visualization: Cluster Profiles

```{r}
#| fig-width: 10
#| fig-height: 6

cluster_means |>
  pivot_longer(-cluster) |>
  ggplot(aes(x = cluster, y = value, fill = name)) +
  geom_col(position = "dodge", width = 0.6) +
  labs(
    title = "Cluster Profiles (K = 3)",
    x = "Cluster",
    y = "Mean Value",
    fill = "Variable"
  ) +
  theme_bw()
```

Because variables are on very different scales, let's also look at standardized profiles:

```{r}
#| fig-width: 10
#| fig-height: 6

mtcars_clustered |>
  mutate(across(where(is.numeric), scale)) |>
  group_by(cluster) |>
  summarize(across(where(is.numeric), mean)) |>
  pivot_longer(-cluster) |>
  ggplot(aes(x = name, y = value, fill = cluster)) +
  geom_col(position = "dodge", width = 0.6) +
  labs(
    title = "Standardized Cluster Profiles (K = 3)",
    x = "Variable",
    y = "Standardized Mean",
    fill = "Cluster"
  ) +
  coord_flip() +
  theme_bw()
```

### Visualization: Scatterplots by Cluster

```{r}
#| fig-width: 8
#| fig-height: 6

mtcars_clustered |>
  ggplot(aes(x = wt, y = mpg, color = cluster)) +
  geom_point(size = 3) +
  labs(
    title = "MPG vs Weight by Cluster",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  ) +
  theme_bw()
```

```{r}
#| fig-width: 8
#| fig-height: 6

mtcars_clustered |>
  ggplot(aes(x = hp, y = mpg, color = cluster)) +
  geom_point(size = 3) +
  labs(
    title = "MPG vs Horsepower by Cluster",
    x = "Horsepower",
    y = "Miles per Gallon"
  ) +
  theme_bw()
```

### Cluster Interpretation

```{r}
cluster_means |>
  select(cluster, mpg, cyl, hp, wt, disp) |>
  knitr::kable(digits = 1, caption = "Key Variables by Cluster")
```

**Cluster descriptions:**

- **Cluster 0**: Economy cars — high mpg, low displacement, fewer cylinders (4-cyl), and lightweight. Examples include cars like the Honda Civic, Toyota Corolla, and Fiat 128.
- **Cluster 1**: Performance / heavy vehicles — low mpg, high horsepower, large displacement (8-cyl), and heavy weight. These include muscle cars and luxury sedans like the Cadillac Fleetwood, Chrysler Imperial, and Camaro Z28.
- **Cluster 2**: Mid-range vehicles — moderate mpg, horsepower, and weight, typically 6-cylinder. Examples include the Mazda RX4, Merc 280, and Hornet 4 Drive.

## Shutdown H2O

```{r}
h2o.shutdown(prompt = FALSE)
```
